\documentclass[11pt]{amsart}
\usepackage[margin=1.2in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[colorlinks=true,linkcolor=blue!60!black,citecolor=blue!60!black,urlcolor=blue!60!black]{hyperref}
\usepackage{xcolor}
\usepackage{booktabs}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{construction}[theorem]{Construction}
\newtheorem{convention}[theorem]{Convention}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{warning}[theorem]{Warning}

\newcommand{\PP}{\mathbf{P}}
\newcommand{\CC}{\mathbf{C}}
\newcommand{\ZZ}{\mathbf{Z}}
\newcommand{\QQ}{\mathbf{Q}}
\newcommand{\Res}{\operatorname{Res}}
\newcommand{\adj}{\operatorname{adj}}
\newcommand{\Pic}{\operatorname{Pic}}
\newcommand{\Br}{\operatorname{Br}}
\newcommand{\divi}{\operatorname{div}}
\newcommand{\Hx}{H_x}
\newcommand{\Hy}{H_y}
\newcommand{\calR}{\mathcal{R}}
\newcommand{\repo}{\url{https://github.com/mattrobball/unirational}}

\title[Unirationality of smooth $(2,3)$ hypersurfaces in $\PP^2\times\PP^2$]
{Unirationality of every smooth $(2,3)$ hypersurface in
$\PP^2\times\PP^2$ by tangent residuals and Latt\`es variation:\\
the construction and its verification artifacts}
\date{July 22, 2026}

\begin{document}

\begin{abstract}
Let $X\subset\PP^2_x\times\PP^2_y$ be any smooth hypersurface of bidegree
$(2,3)$ over $\CC$.  We prove that $X$ is unirational.  A suitable vertical
surface $S=X\cap(\PP^2\times L)$ is rational by Tsen's theorem and maps
birationally, by fiberwise tangent residuals in the plane-cubic fibration,
onto a horizontal divisor $T\sim a\Hx+\Hy$ with $1\leq a\leq10$.  This
gives a dominant rational parametrization of even degree $2a\leq20$.
The key uniform input is the quartic residual-line Latt\`es map of a smooth
plane cubic: its critical-value curve is the dual sextic, so it recovers the
embedded cubic and cannot remain constant in a smooth nonconstant family.
For a general equation a universal resultant identity gives
$T\sim10\Hx+\Hy$ and degree exactly $20$.  Combined with known results, the
construction yields smooth threefolds that are unirational and irrational,
with very general members additionally not stably rational.  These notes are
self-contained on the mathematics and additionally document every machine
computation the proof relies on, with pointers to the certificate files in
the accompanying repository.
\end{abstract}

\maketitle
\tableofcontents

\section{Introduction and statement}\label{sec:intro}

A \emph{standard conic bundle} over $\PP^2$ is a smooth projective threefold
$X$ with a flat morphism $\pi\colon X\to\PP^2$ whose fibers are plane
conics, with relative Picard rank one.  Its discriminant
$\Delta\subset\PP^2$ is a curve; the birational geometry of $X$ is largely
governed by $\deg\Delta$.  Rationality is classical: $X$ is rational iff
$\deg\Delta\le 4$, or $\deg\Delta=5$ with even theta-characteristic
(Beauville, Shokurov; see \cite{Prokhorov} for a survey), while a conic
bundle with $\deg\Delta\ge 6$ is irrational by the Chow/Prym criterion
\cite[Th\'eor\`eme~4.9]{Beauville}.  \emph{Unirationality} is far
more delicate.  The sharpest published statement is
\cite[Corollary~4.9]{Massarenti}: over an algebraically closed field, a
smooth conic bundle $Q^1\to\PP^2$ whose discriminant carries a point of
multiplicity $m_p$ with
\[
\delta\;\le\;m_p+7
\]
is unirational.  The mechanism is projection from that point, which drops
the residual discriminant degree by $m_p$; consequently a \emph{smooth}
discriminant ($m_p=1$) is covered only up to $\delta\le 8$, and degree $9$
only when $\Delta$ is singular.  The unpublished preprint \cite{Mella}
reaches the same boundary in the same shape.  (The frequently quoted bound
$\delta\le7$ of \cite{KollarMella} concerns conic bundle \emph{surfaces},
where $(-K)^2=8-\delta$, not threefolds over $\PP^2$.)  For large general
discriminants the expectation in the literature was negative:
``\emph{It is likely that a conic bundle with sufficiently large and
general discriminant curve is not unirational}''
\cite[\S14.3]{Prokhorov}, echoed in \cite{MassarentiMella}.

Every smooth hypersurface $X$ of bidegree $(2,3)$ in
$\PP^2\times\PP^2$ is a standard conic bundle over the second factor with
discriminant of degree $9$; the general discriminant is smooth.  It
therefore misses the published boundary by exactly one unit.  These notes
give a complete proof of:

\begin{theorem}\label{thm:main}
Every smooth hypersurface $X\subset\PP^2_x\times\PP^2_y$ of bidegree
$(2,3)$ over $\CC$ is unirational.  More precisely, there is a dominant
rational map $\PP^3\dashrightarrow X$ of even degree at most $20$.
\end{theorem}

\begin{theorem}[Generic degree]\label{thm:generic}
For a general $(2,3)$ equation, Construction~\ref{con:map} gives a dominant
rational map $\PP^3\dashrightarrow X$ of degree exactly $20$.
\end{theorem}

\begin{corollary}\label{cor:invisible}
Every smooth $(2,3)$ hypersurface is unirational
{\rm(Theorem~\ref{thm:main})} and irrational
{\rm\cite[Th\'eor\`eme~4.9]{Beauville}}.  A very general member is in
addition not stably rational {\rm\cite[Theorem~1.1]{BvB}}; for such a member,
$\Br(X)=0$ and
$H^3_{\mathrm{nr}}(X,\QQ/\ZZ(2))=0$ {\rm\cite{CTV}}, so the failure of
stable rationality is invisible to the classical unramified invariants;
unlike the Artin--Mumford threefold {\rm\cite{ArtinMumford}}, whose
non-stable-rationality is detected by $\Br$, these examples are detected
only by degeneration of the diagonal {\rm\cite{BvB}}.
\end{corollary}

The construction is elementary in style and could have been carried out
classically; the nearest antecedent is the 1980 note of Del Centina--Verdi
\cite{DCV}, which asserted a \emph{special} smooth unirational $(3,2)$
example by an entirely different method (rationality of one double plane
with special branch sextic).  Their displayed data contain
inconsistencies; a corrected pencil is certified in the repository
(\S\ref{sec:artifacts}).  The fiberwise tangent-residual mechanism used
here appears to be new.

\subsection*{Provenance of these notes}
The construction and proof below were developed in a computational project
archived at \repo.  Its finite algebraic identities were mechanically
verified.  The mathematical exposition is self-contained;
\S\ref{sec:artifacts} is a complete ledger of the machine artifacts the proof
relies on, what each verifies, and which steps rest on standard prose
mathematics instead.

\section{Setup}\label{sec:setup}

\begin{convention}
We work over $\CC$ (any algebraically closed field of characteristic $0$
works verbatim).  Coordinates $x=(x_0{:}x_1{:}x_2)$ on $\PP^2_x$ and
$y=(U{:}V{:}W)$ on $\PP^2_y$.  A bidegree-$(2,3)$ form is written
\[
F(x,y)=x^{\mathsf T}A(y)\,x,
\]
with $A(y)$ a symmetric $3\times3$ matrix of cubic forms in $y$.  We write
$X=\{F=0\}$ and let $\Hx,\Hy$ denote the restrictions of the two hyperplane
classes.  Unless a generic refinement is explicitly stated, $X$ is assumed
smooth.  ``General'' means: in a dense Zariski-open subset of the
parameter space $\PP^{59}=\PP H^0(\mathcal O_{\PP^2\times\PP^2}(2,3))$,
intersected with the smooth locus.
\end{convention}

The threefold $X$ carries \emph{two} fibrations:
\begin{itemize}
\item $\pi=\mathrm{pr}_y\colon X\to\PP^2_y$, a flat conic bundle with
discriminant $\Delta=\{\det A(y)=0\}$ of degree $9$ (smooth for general
$F$);
\item $\rho=\mathrm{pr}_x\colon X\to\PP^2_x$, a fibration whose fiber over
$x$ is the plane cubic $C_x=\{F(x,-)=0\}\subset\PP^2_y$; its generic
fiber is smooth by generic smoothness.
\end{itemize}
The interplay of the two is the whole story.

\begin{lemma}\label{lem:pic}
For smooth $X$: $\Pic(X)=\ZZ \Hx\oplus\ZZ \Hy$ and
$K_X=\mathcal O_X(-1,0)$.  Every effective divisor class $a\Hx+b\Hy$ has
$a,b\ge0$.  Writing $f_y$ (a conic fiber of $\pi$) and $c_x$ (a cubic fiber
of $\rho$) for the two curve classes,
\[
\Hx\cdot f_y=2,\quad \Hy\cdot f_y=0,\qquad
\Hx\cdot c_x=0,\quad \Hy\cdot c_x=3 .
\]
Consequently a surface $S\in|a\Hx+b\Hy|$ has degree $2a$ over $\PP^2_y$ and
degree $3b$ over $\PP^2_x$.  In particular every multisection of the conic
bundle has even degree, and $\rho$ has no rational section.
\end{lemma}

\begin{proof}
Grothendieck--Lefschetz for the ample divisor $X$ in $\PP^2\times\PP^2$
($\dim X=3$) gives $\Pic$; adjunction gives $K_X$.  Effectivity: from
$0\to\mathcal O(a-2,b-3)\to\mathcal O(a,b)\to\mathcal O_X(a,b)\to0$ and
$H^1(\PP^2\times\PP^2,\mathcal O(m,n))=0$ for all $m,n$, one gets
$h^0(\mathcal O_X(a,b))=h^0(\mathcal O(a,b))-h^0(\mathcal O(a-2,b-3))$,
nonzero only if $a,b\ge0$.  The intersection numbers are immediate.  The
parity statement is consistent with, and geometrically explains, the
nontriviality of the $2$-torsion Brauer class of the generic conic
(restriction--corestriction).
\end{proof}

\begin{remark}
Since every surface in the threefold $X$ is a divisor,
Lemma~\ref{lem:pic} constrains all of them.  Adjunction gives
$K_S=((a-1)\Hx+b\Hy)|_S$ for $S\in|a\Hx+b\Hy|$; for a \emph{horizontal}
class ($a\ge1$) this is effective or trivial, so \emph{no smooth member of
any horizontal linear system is rational}.  Any horizontal surface with
rational normalization --- which Theorem~\ref{thm:main} requires, by
Proposition~\ref{prop:engine} below --- is necessarily a singular member of
its system.  This is the precise sense in which the problem resists naive
searches, and why the construction below must produce a highly singular
divisor.
\end{remark}

\begin{proposition}[Multisection principle]\label{prop:engine}
Let $T\subset X$ be an irreducible surface dominating $\PP^2_y$ with
rational normalization $\widetilde T$, of degree $N$ over $\PP^2_y$.  Then
$X$ is unirational of degree $N$.  Conversely, if $X$ is unirational it
contains such surfaces.
\end{proposition}

\begin{proof}
Base-change: $X\times_{\PP^2_y}\widetilde T\to\widetilde T$ is a conic
bundle over a rational surface carrying the tautological section
$t\mapsto(t,t)$.  A conic with a rational point over the field
$\CC(\widetilde T)$ is isomorphic to $\PP^1$, so
$X\times_{\PP^2_y}\widetilde T$ is birational to
$\widetilde T\times\PP^1$, a rational threefold; the projection to $X$ is
dominant of degree $N$.  For the converse, the image of a general plane
under a dominant map $\PP^3\dashrightarrow X$ is a horizontal surface with
rational (indeed unirational, hence by Castelnuovo rational)
normalization.
\end{proof}

Harris's formulation of the problem
--- find rational surfaces in $X$ \emph{other than} the preimages of
rational curves in $\PP^2_y$ \cite[p.~24]{Harris} --- asks exactly for the
horizontality in Proposition~\ref{prop:engine}: the preimage
$\pi^{-1}(\text{curve})$ is vertical and useless.  The point of the
construction below is that one of these ``useless'' vertical surfaces can
be \emph{converted} into a horizontal one using the second fibration
$\rho$.

\section{The generic degree-$20$ construction}\label{sec:construction}

In this section $F$ is general.  The universal identity survives every
specialization; Section~\ref{sec:all-smooth} will use its full line-space
variation to remove this generality assumption.

\begin{construction}\label{con:map}
Fix a general line $L\subset\PP^2_y$ and choose coordinates with
$L=\{W=0\}$.
\begin{enumerate}
\item[\textbf{(I)}] \textbf{The vertical seed.}  Let
$S=X\cap(\PP^2_x\times L)\in|\mathcal O_X(0,1)|$.  Projection to
$L\simeq\PP^1$ makes $S$ a conic bundle over $\PP^1$; its generic fiber is
a smooth conic over the $C_1$ field $\CC(L)\simeq\CC(t)$, hence has a
rational point by Tsen's theorem, and $S$ is a \emph{rational} surface
(irreducible, since the generic fiber is geometrically integral).  Under
the other projection, $S$ is a \emph{trisection} of $\rho$: for general
$x$, the fiber is $C_x\cap L=\{p_1,p_2,p_3\}$.
\item[\textbf{(II)}] \textbf{Tangent residuals.}  For general $x$ the
cubic $C_x$ is smooth at each $p_i$ (this needs only that the binary cubic
$g_x=F(x,-)|_L$ has distinct roots).  Define
\[
\tau\colon S\dashrightarrow X,\qquad
(x,p)\longmapsto\bigl(x,\ \tau_x(p)\bigr),\qquad
T_pC_x\cdot C_x=2p+\tau_x(p).
\]
In the group law of $C_x$ (any flex as origin) this is $p\mapsto -2p$.
\item[\textbf{(III)}] \textbf{The horizontal image.}  Let
$T=\overline{\tau(S)}$.  Theorem~\ref{thm:class} below shows
$T\in|10\Hx+\Hy|$ and $\tau\colon S\dashrightarrow T$ is birational.  By
Lemma~\ref{lem:pic}, $T$ has degree $T\cdot f_y=20$ over $\PP^2_y$; it is
horizontal with rational normalization, so Proposition~\ref{prop:engine}
gives a dominant degree-$20$ map
$\widetilde T\times\PP^1\dashrightarrow X$, proving
Theorem~\ref{thm:generic}.
\end{enumerate}
\end{construction}

Everything therefore rests on identifying the image divisor exactly.  This
is done by a universal polynomial identity.

\subsection{The universal resultant identity}

Write a ternary cubic in the coordinates adapted to $L=\{W=0\}$ as
\begin{equation}\label{eq:f}
f(U,V,W)=g(U,V)+W\,h(U,V)+W^2(iU+jV)+kW^3,
\end{equation}
with $g$ a binary cubic and $h$ a binary quadratic.  For
$p=[u{:}v{:}0]$ with $g(u,v)=0$, the tangent line to $\{f=0\}$ at $p$,
evaluated at $[U{:}V{:}W]$, is
\[
P_{U,V,W}(u,v)=U\,g_u(u,v)+V\,g_v(u,v)+W\,h(u,v),
\]
because $f_U,f_V,f_W$ restrict on $W=0$ to $g_u,g_v,h$.  Set
\[
\calR_f(U,V,W)=\Res_{u,v}\bigl(g,\;P_{U,V,W}\bigr),
\]
the resultant of binary forms of degrees $3$ and $2$; up to the standard
normalization, $\calR_f$ is the product of the three tangent lines at the
points of $\{g=0\}$, a cubic in $(U,V,W)$.

\begin{theorem}[Universal identity]\label{thm:identity}
Let $\Delta(g)$ denote the classical discriminant of the binary cubic
$(18abcd-4b^3d+b^2c^2-4ac^3-27a^2d^2$ for $g=au^3+bu^2v+cuv^2+dv^3)$.
Then, identically in the ten coefficients of $f$,
\begin{equation}\label{eq:identity}
\calR_f+\Delta(g)\,f=W^2\,q_f,
\end{equation}
where $q_f$ is a \emph{linear} form in $(U,V,W)$ whose three coefficients
are homogeneous of degree $5$ in the coefficients of $f$ and have greatest
common divisor $1$ in the universal coefficient ring.
\end{theorem}

\begin{proof}
Expand the left side in $W$.  The $W^0$ term is
$\Res(g,\,Ug_u+Vg_v)+\Delta(g)g(U,V)$.  By Euler's relation
$ug_u+vg_v=3g$, each factor $Ug_u(p_i)+Vg_v(p_i)$ of the resultant is a
linear form vanishing at the root $p_i$, so the product is proportional to
$g(U,V)$; the constant is $-\Delta(g)$ (a classical covariant identity,
verified symbolically in the artifacts).  The $W^1$ coefficient of
$\calR_f$ is a binary quadratic in $(U,V)$; evaluating at a root $p_k$ of
$g$ kills every term of the root-expansion except the one attached to
$p_k$, and comparison at the three roots identifies it with
$-\Delta(g)h(U,V)$ (two binary quadratics agreeing at three distinct
points are equal; the case $\Delta(g)=0$ follows by continuity of a
polynomial identity).  Hence the left side of \eqref{eq:identity} is
divisible by $W^2$, and being cubic, the quotient $q_f$ is linear.  The
coefficient degrees: $\Res_{3,2}$ has bidegree $(2,3)$ in the coefficients
of its two arguments, and the coefficients of $P$ are linear in those of
$f$, giving degree $2+3=5$; likewise $\deg\Delta(g)\cdot f=4+1$.
Primitivity ($\gcd=1$) is a machine fact
(\S\ref{sec:artifacts}, artifact~A1/A4).
\end{proof}

\begin{remark}[Geometric content]\label{rem:geom}
Restricted to the cubic $C=\{f=0\}$, the product of the three tangent
lines has divisor $\sum_i(2p_i+r_i)$ where $r_i=\tau(p_i)$, while $W$ has
divisor $p_1+p_2+p_3$.  On the cubic, $2p_i+r_i\sim \Hy|_C$ and
$\sum p_i\sim \Hy|_C$, whence
\[
r_1+r_2+r_3\;\sim\;3\Hy|_C-2\Hy|_C=\Hy|_C .
\]
A degree-$3$ effective divisor on a smooth plane cubic lying in $|\Hy|_C|$
is a line section: \emph{the three tangent residuals are collinear}, and
$\{q_f=0\}$ is the line through them.  (Group-law formulation: the
residual map is $p\mapsto-2p$ and
$(-2p_1)+(-2p_2)+(-2p_3)=-2(p_1+p_2+p_3)=0$.)
\end{remark}

\subsection{The image divisor and its class}

Now let the coefficients of $f=F(x,-)$ vary: they are \emph{quadratic}
forms in $x$.  Then $\Delta(g_x)$ has bidegree $(8,0)$, $\calR$ has
bidegree $(10,3)$, and the covariant line $q=q_{F(x,-)}(y)$ has bidegree
$(10,1)$ --- linear in $y$, with three coefficient forms
$q_U,q_V,q_W\in\CC[x]_{10}$.

\begin{theorem}\label{thm:class}
For general $F$:
\begin{enumerate}
\item On $X$ the identity \eqref{eq:identity} reads $\calR=W^2q$, giving
the exact divisor equality
\[
\divi_X(\calR)=2S+T_q,\qquad
T_q:=\divi_X(q)\in|10\Hx+\Hy|,\qquad (10,3)=2\,(0,1)+(10,1).
\]
\item $T_q$ has no component vertical over $\PP^2_x$; its fiber over a
general $x$ is exactly the three distinct tangent residuals
$\{r_1,r_2,r_3\}$, reduced.
\item $T_q=T=\overline{\tau(S)}$ is irreducible and
$\tau\colon S\dashrightarrow T$ is birational.
\end{enumerate}
\end{theorem}

\begin{proof}
(1) $q|_X\not\equiv0$ since bidegree $(10,1)$ admits no factor of bidegree
$(2,3)$, and then $\calR|_X=W^2q|_X\not\equiv0$; $W$ cuts $S$ (reduced and
irreducible for general $F,L$).

(2) A component of $T_q$ vertical over $\PP^2_x$ (i.e.\ of class $(a,0)$)
would force $q(x,-)\equiv0$ along a curve of $x$'s, i.e.\ a common factor
of $q_U,q_V,q_W$.  For each potential common-factor degree $e\ge1$, the
locus of coefficient triples with such a factor is the image of the proper
multiplication map
$\PP(\CC[x]_e)\times\PP(\CC[x]_{10-e})^{?}\to\PP(\CC[x]_{10})^{3}$-incidence,
hence closed; there are finitely many $e$; so
$\{F:\gcd(q_U,q_V,q_W)=1\}$ is open.  It is nonempty by an explicit
witness (artifact A1/A4), hence dense.  Then $q(x,-)\equiv 0$ for only
finitely many $x$, and no vertical divisorial component exists.  For the
fiber: over general $x$, $C_x\cap\{q(x,-)=0\}$ is three points by
B\'ezout, and by Remark~\ref{rem:geom} the three residuals $r_i$ lie
there; that they are pairwise distinct and off $L$ for general $x$ is an
open condition on $F$ satisfied at an explicit witness (see
Lemma~\ref{lem:openness} and artifact A1/A4), so the fiber is exactly
$\{r_1,r_2,r_3\}$, reduced.

(3) $\overline{\tau(S)}\subseteq T_q$ by \eqref{eq:identity}; both have
degree $3$ over $\PP^2_x$ ($T_q\cdot c_x=3$ by Lemma~\ref{lem:pic}), and
by (2) the generic fiber of $T_q$ is exactly the injective image of the
generic fiber of $S$.  Hence $T_q$ is the closure of the image of the
irreducible surface $S$ (irreducible), and $\tau$ is generically injective
of degree $1$.
\end{proof}

\begin{lemma}[Openness of the product projection]\label{lem:openness}
Let $\mathcal U\subset\PP^{59}\times\PP^2_x$ be the open locus of pairs
$(F,x)$ where $g_x$ has distinct roots, $q_{F(x,-)}\ne0$, the three
residuals are pairwise distinct, and none lies on $L$.  Then
$\{F:\exists x,\ (F,x)\in\mathcal U\}$ is open in $\PP^{59}$, because the
product projection $\PP^{59}\times\PP^2_x\to\PP^{59}$ is flat and locally
of finite presentation, hence open.  A single explicit pair
$(F_0,x_0)\in\mathcal U$ therefore makes this open nonempty, hence dense.
Concretely, all four
conditions are checked at a witness by: $\operatorname{disc}(g_{x_0})\ne0$;
$q\neq0$; $\operatorname{disc}\bigl(F(x_0,-)|_{\{q=0\}}\bigr)\ne0$ (three
distinct residuals, no root extraction needed); and
$F(x_0,-)\ne0$ at the point $\{q=0\}\cap L$.
\qed
\end{lemma}

\begin{remark}[Independent Chow-theoretic check of the class]
In $\PP^2_x\times L_p\times\PP^2_r$ with hyperplane classes $H,P,R$, the
incidence $\{p\in C_x\cap L,\ r\in T_pC_x\cap C_x\}$ is a complete
intersection of classes $(2H{+}3P)$, $(2H{+}2P{+}R)$, $(2H{+}3R)$; pushing
down $L_p$ gives $20H^2+36HR+9R^2$.  Subtracting twice the pushed diagonal
(tangency multiplicity $2$) leaves
\[
20H^2+32HR+3R^2=(2H+3R)(10H+R),
\]
i.e.\ the residual cycle is a divisor of class $(10,1)$ on the
$(2,3)$ hypersurface --- the same answer by an independent route.
\end{remark}

\subsection{The explicit parametrization}\label{sec:explicit}

For computations (and for the external verification of
\S\ref{sec:artifacts}) it is convenient to have the composite map in
closed staged form.  With $L=\{W=0\}$, parametrize $L$ by
$y(t)=(t{:}1{:}0)$, and suppose (after the harmless linear specialization
described in \S\ref{sec:artifacts}, A4) that $e=(1{:}0{:}0)$ lies on every
conic $F(-,y(t))$.  Writing $A_{ij}(t)$ for the restricted coefficients of
the conic:
\begin{align*}
\textbf{(I)}\quad
&x^*(t,m)=\bigl(A_{11}+mA_{12}+m^2A_{22},\\
&\hspace{7em}-(A_{01}+mA_{02}),\ -m(A_{01}+mA_{02})\bigr)
\end{align*}
This is the second point where the line through $e$ in direction $(0,1,m)$
meets the conic.  Hence it birationally parametrizes $S$ by $(t,m)$.
\begin{align*}
\textbf{(II)}\quad
&f^*=F(x^*(t,m),-),\qquad
(\alpha,\beta,\gamma)=\nabla_y f^*\big|_{y(t)},\qquad
z^*=(\gamma,0,-\alpha),\\
&f^*\bigl(s\,y(t)+u\,z^*\bigr)=u^2\bigl(\mathsf A\,s+\mathsf B\,u\bigr),
\qquad
r(t,m)=\mathsf B\,y(t)-\mathsf A\,z^*
\end{align*}
($z^*$ is a second point of the tangent line, since
$\alpha\gamma+\gamma(-\alpha)=0$; the two leading coefficients vanish
identically, expressing tangency, and $r$ is the residual).
\begin{align*}
\textbf{(III)}\quad
&d=(0,1,n),\qquad Q=F(-,r),\\
&B_Q(u,v)=Q(u+v)-Q(u)-Q(v),\\
&x'(t,m,n)=Q(d)\,x^*-B_Q(x^*,d)\,d .
\end{align*}
Then $\Phi(t,m,n)=(x'(t,m,n);\,r(t,m))$ is a dominant rational map
$\mathbf A^3\dashrightarrow X$ of degree $20$.  Stage (III) is the chord
construction on the conic fiber through the marked point $x^*$, and its
correctness is the one-line \emph{reflection identity} for any ternary
quadratic form $Q$ with polarization $B_Q$:
\[
Q\bigl(Q(d)\,x-B_Q(x,d)\,d\bigr)=Q(d)^2\,Q(x),
\]
so $F(x',r)=Q(d)^2F(x^*,r)=0$.
The degree over $X$ equals $\deg(T/\PP^2_y)=T\cdot f_y=20$.

\begin{remark}[Scope]\label{rem:scope}
The construction uses the plane-cubic fibration $\rho$, which exists
precisely because the second degree is $3$.  It therefore says nothing
about arbitrary standard conic bundles with smooth degree-$9$ discriminant
(the matrix entries need not be all cubic), nor about bidegrees $(2,d)$,
$d\ge4$, where the fibers of $\rho$ have genus $>1$ and the tangent line
meets the fiber in $d-2>1$ residual points (a correspondence, not a map).
Harris's question for $d$ large \cite[p.~24]{Harris} remains open.
\end{remark}

\begin{remark}[Byproducts in the repository]
Two independent negative results certified in the same project delimit how
sharp the construction is: on a general $X$, \emph{every} member of
$|\Hx|$ is a $K3$ double plane (its branch sextic never acquires a point
of multiplicity $\ge3$), and \emph{no} member of $|\Hx+\Hy|$ has rational
normalization.  So the minimal-class searches fail provably, and the first
horizontal rational surface found lives in $|10\Hx+\Hy|$ --- a singular
member, as forced by adjunction (Lemma~\ref{lem:pic} ff.).
\end{remark}

\section{Removing the generality hypothesis}\label{sec:all-smooth}

We now prove Theorem~\ref{thm:main}.  The missing uniform point is not the
resultant identity, which is universal, but horizontality: after a special
specialization the three degree-$10$ coefficients of the residual line may
have a common factor, and for a poorly chosen $L$ the residual line could be
constant in $x$.  Varying $L$ in the full dual plane rules this out.

\subsection{The residual-line Latt\`es map}

Let $C\subset\PP^2$ be a smooth plane cubic.  We work over a field of
characteristic zero.  Write
\[
T_pC\cdot C=2p+g(p).
\]
After choosing a flex as origin, $g(p)=-2p$.  Thus $g$ is a finite morphism
of degree $4$ and preserves collinearity.  If $L\cdot C=p+q+r$, the three
points $g(p),g(q),g(r)$ are again collinear.  They define a morphism
\[
\delta_C:(\PP^2)^\vee\longrightarrow(\PP^2)^\vee,
\qquad L\longmapsto\overline{g(p)g(q)g(r)}.
\tag{4.1}
\]

\begin{lemma}[Critical values]\label{lem:lattes}
The morphism $\delta_C$ has algebraic degree $4$, and its reduced
critical-value curve is the dual sextic $C^\vee$.  In particular
$\delta_C$ determines the embedded cubic $C$ by biduality.
\end{lemma}

\begin{proof}
Over an algebraic closure choose a flex as origin.  The morphism
\[
\pi:C\times C\longrightarrow(\PP^2)^\vee,
\qquad(p,q)\longmapsto\overline{pq},
\]
with the tangent line as its value on the diagonal, is the quotient by the
natural $S_3$-action on the triple $(p,q,-p-q)$.  Its ramification divisor
is the union of the three reflection curves on which two entries coincide,
and its branch curve is $C^\vee$.  Put $\mu=[-2]\times[-2]$.  Then
\[
\pi\mu=\delta_C\pi,
\qquad
\pi^*R_{\delta_C}=\mu^*R_\pi-R_\pi
=\sum_{i<j}\ \sum_{0\ne\varepsilon\in C[2]}
\{p_i-p_j=\varepsilon\}.
\tag{4.2}
\]
The curves on the right map to, and sweep out, the locus where two residual
points coincide, namely $C^\vee$.  This proves the critical-value assertion.
Since $[-2]^*\mathcal O_C(1)$ has degree four times that of
$\mathcal O_C(1)$, the induced plane map has algebraic degree $4$.
Finally $(C^\vee)^\vee=C$.

This construction and the critical-value statement also appear explicitly
in \cite[p.~964, (2.6)]{MMW}; see also \cite[\S4.3]{DabijaJonsson}.
The quotient proof above is algebraic and therefore applies over any
characteristic-zero extension field.
\end{proof}

\begin{lemma}[Constant-value descent]\label{lem:constant-descent}
Let $K=\CC(s,t)$ and
$\varphi:\PP^n_K\to\PP^m_K$ be a morphism.  If
$\varphi(z)\in\PP^m(\CC)$ for a Zariski-dense set of
$z\in\PP^n(\CC)$, then $\varphi$ is defined over $\CC$.
\end{lemma}

\begin{proof}
Pass to source and target affine charts retaining a dense subset.  Write the
coordinate ratios as $R_i(z)\in K(z_1,\ldots,z_n)$.  Applying either
$\CC$-derivation of $K$ shows that $D(R_i)$ vanishes at a dense set of
constant points, hence vanishes identically.  The common constants of these
derivations in $K(z_1,\ldots,z_n)$ are
$\CC(z_1,\ldots,z_n)$, so all ratios descend.  The descended coordinate
forms have no common zero because their base change is the given morphism.
\end{proof}

\subsection{Choosing a horizontal line section}

Let $X$ be an arbitrary smooth $(2,3)$ hypersurface and put
\[
K=\CC(\PP^2_x),\qquad C=X_K\subset\PP^2_{y,K}.
\]
The generic cubic $C$ is smooth by generic smoothness.  For completeness,
smoothness of $X$ rules out a fiber equal to the whole $\PP^2_y$: at such a
fiber the two derivatives in independent local $x$-directions are plane
cubics and have a common zero, which would be singular on $X$.

The embedded cubic $C$ is not constant.  Otherwise its equation over $K$
would be $\lambda(x)f_0(y)$ for a fixed cubic $f_0$.  Comparing one nonzero
coefficient makes $\lambda$ a quadratic form, so
$F(x,y)=Q(x)f_0(y)$ in $\CC[x,y]$, contradicting smoothness.

Apply Lemma~\ref{lem:lattes} to $C/K$.  The morphism $\delta_C$ cannot be
defined over $\CC$: if it were, its critical-value sextic and then its
bidual $C$ would be constant.  By Lemma~\ref{lem:constant-descent}, the
constant lines having constant image under $\delta_C$ are not Zariski dense.
We may therefore choose a constant line $L\subset\PP^2_y$ such that
\[
\delta_C(L)\notin(\PP^2_y)^\vee(\CC),
\tag{4.3}
\]
and simultaneously impose the following open conditions: $L$ is not
contained in the conic discriminant, $C\cap L$ is reduced, and $[-2]$ is
injective on its three points.  The last bad locus is, after algebraic
closure, the finite union of curves swept out by the chords
$\overline{p,p+\varepsilon}$ for $0\ne\varepsilon\in C[2]$; the union is
Galois invariant.  A proper closed subset of the constant dual plane
base-changed to $K$, together with these proper $K$-closed loci, cannot
contain every constant point (expand its finitely many $K$-coefficients in a
finite $\CC$-linearly independent set).  Hence such an $L$ exists.

The surface
\[
S_L=X\cap(\PP^2_x\times L)
\]
is integral: its generic fiber over $L$ is a geometrically integral conic,
and smoothness of $X$ also rules out a whole $\PP^2_x$ fiber by the same
two-derivative argument.  Tsen's theorem then gives
$\CC(S_L)=\CC(t)(z)$, so $S_L$ is rational.  Tangent residuals define
$\tau_L:S_L\dashrightarrow X$.  The chosen injectivity condition shows
that its image $T_L$ is irreducible, $S_L\dashrightarrow T_L$ is birational,
and $T_L\to\PP^2_x$ has degree $3$.

Suppose $T_L$ were not horizontal over $\PP^2_y$.  Its image cannot be a
point, since every conic fiber is one-dimensional.  If its image were a
curve, write $[T_L]=a\Hx+b\Hy$ using Lemma~\ref{lem:pic}.  A general conic
fiber outside that curve is disjoint from $T_L$, so $2a=0$; degree $3$ over
$\PP^2_x$ gives $3b=3$.  Thus $[T_L]=\Hy$.  The restriction sequence gives
\[
H^0(X,\mathcal O_X(0,1))
=H^0(\PP^2_y,\mathcal O(1)),
\]
so $T_L$ would be the inverse image of a constant line $M$.  On the generic
cubic this says $\delta_C(L)=M$, contradicting (4.3).  Hence $T_L$ is
horizontal.

Finally put $L=\{W=0\}$ and use Theorem~\ref{thm:identity}.  The residual
line has three degree-$10$ coefficients in $x$.  After dividing their gcd
of degree $e$, its primitive equation cuts a divisor
\[
D\sim(10-e)\Hx+\Hy
\]
containing $T_L$ as its unique component dominating $\PP^2_x$.  We have
$[T_L]=a\Hx+\Hy$.  Every effective prime divisor
$A\Hx+B\Hy$ has $A,B\geq0$, by intersection with general conic and cubic
fibers.  Comparing coefficients in $D$ therefore gives
\[
1\leq a\leq10-e\leq10.
\]
The normalization of $T_L$ is rational and its degree over $\PP^2_y$ is
$2a\leq20$.  Base change of the conic bundle to this normalization has the
tautological point and is birational to $\widetilde T_L\times\PP^1$.
Projection back to $X$ proves Theorem~\ref{thm:main}.

\begin{remark}[A broader rational case]
If an integral $(2,3)$ member has geometrically integral but singular
generic cubic over $\CC(\PP^2_x)$, its unique geometric singular point is
rational over that field.  Projection from it makes the generic cubic
rational, so $\CC(X)=\CC(\PP^2_x)(t)$ and $X$ itself is rational.  Thus the
smooth generic cubic is the hard case.  Generically geometrically reducible
or nonreduced cubic fibrations require separate arguments.
\end{remark}

\section{Verification artifacts}\label{sec:artifacts}

The all-smooth proof uses one machine-checked algebraic fact, the universal
identity (Theorem~\ref{thm:identity}); its remaining inputs are the prose
Latt\`es, descent, and Picard arguments of Section~\ref{sec:all-smooth}.
The sharper generic degree-$20$ statement additionally uses primitivity of
$q$ and the explicit witnesses of Lemma~\ref{lem:openness}.  These finite
facts are machine-checked.
This section is a complete ledger.  All paths are relative to the
repository \repo; logs of every run are committed beside the scripts.

\subsection*{A0 --- the all-smooth certificate}
The complete standalone proof, including the quotient ramification formula,
constant-value descent, simultaneous choice of $L$, and degree bound, is
\path{certificates/all_smooth_tangent_residual_theorem.md}.

\subsection*{A1 --- the certificate of record}\mbox{}\par
\noindent
\path{problems/B-conic-bundle-multisections/certificates/}\\
\path{tangent_residual_local_checks.py} (with \texttt{.log}):
\begin{itemize}
\item verifies \eqref{eq:identity} \emph{fully symbolically} over the
universal ring $\ZZ[a_3,\dots,k]$ in ten independent coefficient symbols
(the $W^0,W^1$ coefficients of $\calR+\Delta f$ vanish identically; the
quotient by $W^2$ is linear);
\item verifies each coefficient of $q_f$ is homogeneous of degree $5$ and
that the three have $\gcd 1$ universally, and again after a recorded
substitution by ten quadratic forms (so the general $(2,3)$ equation has
primitive $q$);
\item supplies the smooth-fiber witness of Lemma~\ref{lem:openness}:
an explicit $(F_0,x_0)$ with the cubic smooth (unit Gr\"obner bases in
three charts), three residual points computed, verified to lie on the
cubic, on the tangent lines, on $\{q=0\}$, off $L$, pairwise distinct;
\item records the divisor arithmetic $(10,3)-2(0,1)=(10,1)$, degree $20$.
\end{itemize}
The full prose proof of the generic refinement is
\path{certificates/tangent_residual_theorem.md}.

\subsection*{A2 --- independent re-derivation (audit)}
During the audit of this session, the identity was re-derived from scratch
in a fresh SymPy session sharing no code with A1: the $W^0$ divisibility,
the cofactor being the \emph{classical} discriminant (both sign
conventions tested), degree-$5$ homogeneity, and universal $\gcd=1$ were
reconfirmed independently.
% (transcript: session audit, 2026-07-22)

\subsection*{A3 --- independent end-to-end geometric check (audit)}
On an explicit cubic with rational points on $L$
($g=(u-v)(u-2v)(u+v)$, generic $h,i,j,k$), the three tangent residuals
were computed \emph{from the raw geometry} (tangent lines; factoring the
double root out of the restricted cubic) and verified to be distinct, off
$L$, and to lie on $\{q_f=0\}$ --- confirming the geometric
interpretation, not merely the algebra.  A second check drew a random
$(2,3)$ form over $\QQ$ and verified the witness conditions of
Lemma~\ref{lem:openness} at $x_0=(1,2,3)$, plus $\gcd(q_U,q_V,q_W)=1$ by
restriction to two lines.

\subsection*{A4 --- external random-coefficient instance}\mbox{}\par
\noindent
\path{problems/B-conic-bundle-multisections/external_check/}: a full
end-to-end falsification test on coefficients drawn from
\texttt{/dev/urandom} (recorded), with four coefficients zeroed so that
the section $e=(1{:}0{:}0)$ exists over $\QQ$ and the composite map has
exact rational coefficients (a codimension-$4$ linear slice needed only
because Tsen's theorem is not effective over $\QQ$; over $\CC$ the theorem
needs no such conditioning).  The harness \texttt{build\_explicit\_map.py}
constructs the staged map of \S\ref{sec:explicit} and verifies, exactly:
\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
check & statement & method\\
\midrule
C1 & $F(x^*,y(t))\equiv0$ & symbolic\\
C2 & tangency (two leading coefficients vanish) & symbolic\\
C3 & $f^*(r)\equiv0$ & full $100\times53$ grid$^{\dagger}$\\
C4 & $q_{f^*}(r)\equiv0$ & full $62\times37$ grid$^{\dagger}$\\
C5 & reflection identity & generic symbols\\
C6 & $F(\Phi)=0$ at sample rational points & exact evaluation\\
C7 & Jacobian rank $3$ (dominance) & exact symbolic differentiation\\
C8 & fiber over random $y_0$: squarefree of degree $20$ & resultant\\
\bottomrule
\end{tabular}
\end{center}
{\small $^{\dagger}$Deterministic proofs: a polynomial of bidegree at most
$(D_1,D_2)$ vanishing at a $(D_1{+}1)\times(D_2{+}1)$ grid of distinct
rational nodes is zero; the bidegree bounds $(99,52)$ and $(61,36)$ are
derived in the harness.}\\[2pt]
All eight pass (\path{run_log.txt}); the explicit map instance is
\path{map_instance.txt}, and \path{README_CHECK.md} gives
CAS-agnostic instructions for third-party verification with no trust in
the pipeline.

\subsection*{A5 --- unconditioned instance (no slice)}
\path{external_check/unconditioned_check.py} (with \texttt{.log})
addresses the natural objection that the codimension-$4$ slice in A4 might
matter: it runs the full mechanism on a fresh entropy draw with \emph{no}
coefficient zeroed.  With no rational point available, it works in the
cubic \'etale algebra $K=\QQ[\theta]/(g_0)$, $g_0=F(x_0,-)|_L$, where the
three marked points live; reduction modulo the full cubic verifies each
identity at all three conjugate points simultaneously.  It checks:
primitivity of $q_F$; the witness conditions of Lemma~\ref{lem:openness};
tangency, $F(x_0,r)=0$, $q(r)=0$, and $r\notin L$ (via
$\Res(g_0,r_W)\ne0$) in $K$; the stage-(III) point $F(x',r)=0$ in $K$; and
the squarefree degree-$20$ fiber.  All pass.  The slice in A4 is thereby
confirmed to play no role beyond supplying $\QQ$-rationality of the
composite parametrization (Tsen's theorem being non-effective over
$\QQ$); by $\mathrm{PGL}_3\times\mathrm{PGL}_3$-equivariance the sliced
family $\{X\supseteq\{e\}\times L\}$ is in no special position for
anything the proof uses.

\subsection*{Prose-only steps}
For completeness: the steps of the proof resting on standard mathematics
rather than computation are Tsen's theorem, generic smoothness,
Grothendieck--Lefschetz (Lemma~\ref{lem:pic}), the group-law and Latt\`es
quotient arguments, constant-value descent, openness of the flat product
projection in Lemma~\ref{lem:openness}, and the Chow-theoretic cross-check.
Each is supplied in the proof or cited to a primary source.

\subsection*{Independence and audit trail}
A1 was rerun in a clean environment during the audit (Macaulay2 1.26.06 /
SymPy 1.14.0) with output matching the committed logs; A2--A3 share no
code with A1; A4 takes its input from an external entropy source.  The
literature verification was performed against primary sources on
2026-07-22 and re-confirmed on 2026-07-23 by reading the relevant papers
directly.  Specifically checked: \cite[Corollary~4.9]{Massarenti} and its
definition $\delta:=\deg D$, which places the published boundary for a
\emph{smooth} discriminant at $\delta\le8$ and leaves our $\delta=9$
uncovered by exactly one unit; \cite[Corollary~4.10]{Massarenti}
($\delta\le12$ for quadric surface bundles over $\PP^2$, which subsumes the
$\PP^3\times\PP^2$ analogue of Theorem~\ref{thm:main});
\cite[Remark~2.9]{Massarenti} for Lang's theorem in the form
$h>2^{r+n-h}-2$; \cite[Theorem~1.1]{BvB}; \cite[Corollary~1.2]{Mella};
the method and status of \cite{DCV}; and the residual-line Latt\`es
statements in \cite{MMW,DabijaJonsson}.  One caution recorded for future
searches: the bidegree-$(2,3)$ hypersurfaces in $\PP^2\times\PP^3$ of
Auel--Böhning--Pirutka are \emph{cubic surface bundles} over $\PP^2$, a
different family from the one studied here.

\begin{thebibliography}{99}

\bibitem{ArtinMumford}
M.~Artin, D.~Mumford,
\emph{Some elementary examples of unirational varieties which are not
rational}, Proc.\ London Math.\ Soc.\ (3) \textbf{25} (1972), 75--95.

\bibitem{Beauville}
A.~Beauville,
\emph{Vari\'et\'es de Prym et jacobiennes interm\'ediaires},
Ann.\ Sci.\ \'Ecole Norm.\ Sup.\ (4) \textbf{10} (1977), 309--391.

\bibitem{BvB}
C.~B\"ohning, H.-C.~Graf von Bothmer,
\emph{On stable rationality of some conic bundles and moduli spaces of
Prym curves}, Comment.\ Math.\ Helv.\ \textbf{93} (2018), 133--155.

\bibitem{CGM}
L.~Casarotti, S.~Gammelgaard, A.~Massarenti,
\emph{On the unirationality of conic bundles with discriminant of degree
eight}, preprint, \href{https://arxiv.org/abs/2511.17213}{arXiv:2511.17213}
(2025).

\bibitem{CTV}
J.-L.~Colliot-Th\'el\`ene, C.~Voisin,
\emph{Cohomologie non ramifi\'ee et conjecture de Hodge enti\`ere},
Duke Math.\ J.\ \textbf{161} (2012), 735--801.

\bibitem{DCV}
A.~Del Centina, L.~Verdi,
\emph{Su certi fibrati in coniche di $\PP^2\times\PP^2$, alcuni dei quali
unirazionali e non razionali},
Atti Accad.\ Naz.\ Lincei Rend.\ Cl.\ Sci.\ Fis.\ Mat.\ Nat.\ (8)
\textbf{69} (1980), no.~6, 338--345.

\bibitem{DabijaJonsson}
M.~Dabija, M.~Jonsson,
\emph{Algebraic webs invariant under endomorphisms},
Publ.\ Mat.\ \textbf{54} (2010), no.~1, 137--148,
\href{https://doi.org/10.5565/PUBLMAT_54110_07}{doi:10.5565/PUBLMAT\_54110\_07}.

\bibitem{Harris}
J.~Harris,
\emph{Rationality, unirationality and rational connectivity},
lecture notes, AMS Summer Institute in Algebraic Geometry (Seattle),
\url{https://www.math.columbia.edu/~thaddeus/seattle/harris1.pdf}.

\bibitem{KollarMella}
J.~Koll\'ar, M.~Mella,
\emph{Quadratic families of elliptic curves and unirationality of degree
1 conic bundles}, Amer.\ J.\ Math.\ \textbf{139} (2017), 915--936.

\bibitem{Massarenti}
A.~Massarenti,
\emph{On the unirationality of quadric bundles},
Adv. Math. \textbf{431} (2023), 109235;
\href{https://arxiv.org/abs/2204.08793}{arXiv:2204.08793}.

\bibitem{MassarentiMella}
A.~Massarenti, M.~Mella,
\emph{On the birational geometry of conic bundles over the projective
space}, Math.\ Nachr.\ \textbf{297} (2024), 1208--1220.

\bibitem{Mella}
M.~Mella,
\emph{On the unirationality of 3-fold conic bundles}, preprint,
\href{https://arxiv.org/abs/1403.7055}{arXiv:1403.7055} (2014).

\bibitem{MMW}
C.~T.~McMullen, R.~E.~Mukamel, A.~Wright,
\emph{Cubic curves and totally geodesic subvarieties of moduli space},
Ann.\ of Math.\ (2) \textbf{185} (2017), no.~3, 957--990,
\href{https://doi.org/10.4007/annals.2017.185.3.6}{doi:10.4007/annals.2017.185.3.6}.

\bibitem{Prokhorov}
Yu.~Prokhorov,
\emph{The rationality problem for conic bundles},
Russian Math.\ Surveys \textbf{73} (2018), no.~3, 375--456.

\end{thebibliography}

\end{document}
